---
title: "Data plan for geneExp"
author: "Bingfang"
date: "12/12/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Objective

Identify a list of true expressed genes of a specific cell type, which will use to for functional analysis of biological process/pathway of the cell type.

## 2. Set environment


```{r, echo=TRUE,warning=FALSE,message=FALSE}
#install.packages("sads")  # not working on local MAC
library(sads)
#install.packages("BiocParallel")
library(BiocParallel)
library(Seurat)
library(org.Hs.eg.db)
#install.packages("Matrix")
library(Matrix)

# Get gene-specific distributions
load('params_EM_81020.rda')
pi.all.g <- params[[1]]
g.on.all.g <- params[[2]]
g.off.all.g <- params[[3]]
a.all.g <- params[[4]]
sigma.all.g <- params[[5]]
mu.g<-c(-12.73153,-8.301349)
```

## 2. Data collection and preprocess.

### 2.1 Read .rds file

HLCA data from cellXgene

cellRef data from cellXgene

```{r} 

hlca = readRDS("/gpfs/gsfs10/users/xubr/cellXgene/scRNAseq-cell-type/data/b351804c-293e-4aeb-9c4c-043db67f4540_hlca_core.rds")
#cellRef = readRDS("/gpfs/gsfs10/users/xubr/cellXgene/scRNAseq-cell-type/data/

```

### 2.2 remove low quality cells

```{}
hlca_filtered <- subset(hlca,subset = nFeaturess_RNA_detected > 200 & nFeaturess_RNA_detected < 2500)

## need to convert to gene symbol to filter percent.mt < 5
##hlca_filtered <- subset(hlca,subset = nFeaturess_RNA_detected > 200 & nFeaturess_RNA_detected < 2500 & percent.mt < 5)
```

### 2.3 subset by cell type

```{r}
unique(hlca$cell_type)

hlca_mucus_secreting <- subset(hlca, subset = cell_type == "mucus secreting cell")
hlca_ionocyte <- subset(hlca, subset = cell_type == "ionocyte")
hlca_2types <- subset(hlca, subset = cell_type %in% c("mucus secreting cell", "ionocyte"))


typeof(hlca_mucus_secreting)
names(hlca_mucus_secreting)
head(hlca_mucus_secreting$RNA)


typeof(hlca_mucus_secreting$RNA)
dim(hlca_mucus_secreting$RNA)
head(hlca_mucus_secreting$RNA@counts)


```


## 2.4 geneXcell matrix


```{r}
# hlca  is seurat object from readRDS
sparse_matrix_mucus_secreting <- hlca_mucus_secreting@assays$RNA@counts
hlca_matrix_mucus_secreting <- as.matrix(sparse_matrix_mucus_secreting) ## if use full dataset, it will generate 130G matrix

sparse_matrix_ionocyte <- hlca_ionocyte@assays$RNA@counts
hlca_matrix_ionocyte <- as.matrix(sparse_matrix_ionocyte) 

sparse_matrix_2types<-hlca_2types@assays$RNA@counts
hlca_matrix_2types <- as.matrix(sparse_matrix_2types) 
hlca_matrix_2types_labels<-c(rep('mucus_secreting',dim(hlca_matrix_mucus_secreting)[2]),rep('ionocyte',dim(hlca_matrix_ionocyte)[2]))

dim(hlca_matrix_2types)                            
length(hlca_matrix_2types_labels)
sum(is.na(hlca_matrix_2types))

datSum<-rowSums(hlca_matrix_2types)
datSum[1:10]
```

## 4. calculate rate and probability



```{r, message=FALSE}

# Train reference data
trainReference <- function(ref,pi.all=pi.all.g,mu=mu.g,a=a.all.g,sigma.all=sigma.all.g,
                           g.on.all=g.on.all.g,g.off.all=g.off.all.g,discrim_only=FALSE) {
  # Subset to genes present in both reference data and knowledge base
  common <- intersect(names(ref),rownames(pi.all))
  if (discrim_only) {
    common <- intersect(common,discrim.g)
  }
  N <- sum(ref)
  ref <- ref[common]
  pi.all2 <- pi.all[common,]
  a.all2 <- a[common]
  sigma.all2 <- sigma.all[common,]
  g.on.all2 <- g.on.all[common]
  g.off.all2 <- g.off.all[common]
  
  # Compute mixing probability for each component
  prob.exp <- pi.all2[,1]*dnbinom(ref,1,1/((N/(a.all2))+1))
  prob.ln1 <- sapply(common,function(j) 
    pi.all2[j,2]*sads::dpoilog(ref[j],mu[1]+g.off.all2[j]+log(N),sigma.all2[j,1]))
  prob.ln2 <- sapply(common,function(j) 
    (1-pi.all2[j,1]-pi.all2[j,2])*sads::dpoilog(ref[j],mu[2]+g.on.all2[j]+log(N),sigma.all2[j,2])) 
  return(data.frame(rate=ref/N,exp=prob.exp/(prob.exp+prob.ln1+prob.ln2),ln1=prob.ln1/(prob.exp+prob.ln1+prob.ln2),ln2=prob.ln2/(prob.exp+prob.ln1+prob.ln2)))
}


# Train all reference
trainAllReference <- function(data,labels,pi.all=pi.all.g,mu=mu.g,a=a.all.g,sigma.all=sigma.all.g,
                              g.on.all=g.on.all.g,g.off.all=g.off.all.g,discrim_only=FALSE) {
  cell.types <- unique(labels)
  dataSums <- lapply(cell.types,function(c) rowSums(data[,labels==c]))

  d.list <- bplapply(dataSums,trainReference,pi.all=pi.all,mu=mu,a=a,sigma.all=sigma.all,
                     g.on.all=g.on.all,g.off.all=g.off.all,discrim_only=discrim_only)
  names(d.list) <- cell.types
  return(d.list)
}


hlca_2types.d <- trainAllReference(hlca_matrix_2types,hlca_matrix_2types_labels,discrim_only=FALSE)
hlca_2types.d[[1]]
```

## 5. getBarcode

```{r}
getBarcode<-function(d.list) {
  barcode <- sapply(d.list,function(x) 1-x[,2]-x[,3])      # ### Barcode is   ln2=prob.ln2/(prob.exp+prob.ln1+prob.ln2) if greater than 0
  barcode[barcode<0] <- 0
  colnames(barcode) <- names(d.list)
  rownames(barcode) <- rownames(d.list[[1]])
  return(barcode)
}

barcodes <- getBarcode(hlca_2types.d)
head(barcodes)

```


## 6. Visualization



## 7. Validation

Compare results of the same cell types from different datasets e.g. cellRef vs. Hlca

Compare results from different cell types. e.g. mucus_secreting cell vs. ionocyte.

Subsampling, Bootstrap Aggregation,Perturbation ( add 10% cells from other cell type)

Compare with the known go term of the cell type.

## 8. Interpretation

## 9. Reporting
Summarize findings in a detailed report or manuscript with figures and statistical analyses.
