---
title: "DataAnalysis plan for geneExp"
author: "Bingfang"
date: "12/12/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Objective

Identify a list of true expressed genes of a specific cell type, which will use to for functional analysis of biological process/pathway of the cell type.

## 2. Set environment


```{r,, echo=TRUE,warning=FALSE,message=FALSE, results=FALSE}
#install.packages("sads")     # not working on local MAC
library(sads)
#install.packages("BiocParallel")
library(BiocParallel)
library(Seurat)                #cannot install on local MAC
library(org.Hs.eg.db)
#install.packages("Matrix")
library(Matrix)

# Get gene-specific distributions
load('params_EM_81020.rda')
pi.all.g <- params[[1]]
g.on.all.g <- params[[2]]
g.off.all.g <- params[[3]]
a.all.g <- params[[4]]
sigma.all.g <- params[[5]]
mu.g<-c(-12.73153,-8.301349)
```

## 2. Data Collection

Input Data:

PBMC demo data 


## 3. Data Preprocessing

PBMC data are raw counts after removing low quality cells. Default filtering was used to removed low quality cells.

```{r}
# Remove low-quality cells 
#pbmc <- subset(pbmc,subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
```

pbmc demo data

```{r,echo=TRUE,warning=FALSE,message=FALSE}
set.seed(6619)

### Preparation for symbol->ENSEMBL conversion
Hs_symbol <- org.Hs.egSYMBOL
mapped_Hs_genes.symbol <- mappedkeys(Hs_symbol)
Hs_symbol.df <- as.data.frame(Hs_symbol[mapped_Hs_genes.symbol])
Hs_ensembl <- org.Hs.egENSEMBL
mapped_Hs_genes.ensembl <- mappedkeys(Hs_ensembl)
Hs_ensembl.df <- as.data.frame(Hs_ensembl[mapped_Hs_genes.ensembl])
Hs_mapping <- merge(Hs_symbol.df,Hs_ensembl.df)

## CD4: https://support.10xgenomics.com/single-cell-gene-expression/datasets/1.1.0/cd4_t_helper? 
cd4_facs.data <- Read10X(data.dir = "./data/cd4_singlecell/hg19/") 
cd4_facs.data <- as.matrix(cd4_facs.data)
rownames(cd4_facs.data) <- Hs_mapping$ensembl_id[match(rownames(cd4_facs.data),
                                                       Hs_mapping$symbol)]
cd4_facs.data <- cd4_facs.data[!is.na(rownames(cd4_facs.data)),]
cd4_facs.data <- na.omit(cd4_facs.data)
cd4.test <- cd4_facs.data[,1:100] # CD4 withheld cells
cd4_facs.data <- cd4_facs.data[,101:5101] # CD4 training cells

## CD8: https://support.10xgenomics.com/single-cell-gene-expression/datasets/1.1.0/cytotoxic_t
cd8_facs.data <- Read10X(data.dir = './data/filtered_matrices_cd8/hg19/')
cd8_facs.data <- as.matrix(cd8_facs.data) 
rownames(cd8_facs.data) <- Hs_mapping$ensembl_id[match(rownames(cd8_facs.data),
                                                       Hs_mapping$symbol)]
cd8_facs.data <- cd8_facs.data[!is.na(rownames(cd8_facs.data)),]
cd8.test <- cd8_facs.data[,1:100] # CD8 withheld cells
cd8_facs.data <- cd8_facs.data[,101:5101] # CD8 training cells

## CD14: https://support.10xgenomics.com/single-cell-gene-expression/datasets/1.1.0/cd14_monocytes
cd14_facs.data <- Read10X(data.dir = './data/filtered_matrices_cd14/hg19/')
cd14_facs.data <- as.matrix(cd14_facs.data)
rownames(cd14_facs.data) <- Hs_mapping$ensembl_id[match(rownames(cd14_facs.data),
                                                        Hs_mapping$symbol)]
cd14_facs.data <- cd14_facs.data[!is.na(rownames(cd14_facs.data)),]
cd14.test <- cd14_facs.data[,1:100] # CD14 withheld cells
cd14_facs.data <- cd14_facs.data[,101:ncol(cd14_facs.data)] # CD14 training cells

## NK: https://support.10xgenomics.com/single-cell-gene-expression/datasets/1.1.0/cd56_nk
nk_facs.data <- Read10X(data.dir = './data/filtered_matrices_nk/hg19/')
nk_facs.data <- as.matrix(nk_facs.data)
rownames(nk_facs.data) <- Hs_mapping$ensembl_id[match(rownames(nk_facs.data),
                                                      Hs_mapping$symbol)]
nk_facs.data <- nk_facs.data[!is.na(rownames(nk_facs.data)),]
nk.test <- nk_facs.data[,1:100] # NK withheld cells
nk_facs.data <- nk_facs.data[,101:5101] # NK training cells
```


```{r,echo=TRUE,warning=FALSE,message=FALSE}
## Input a single matrix with a vector of labels 
common_pbmcs_genes <- Reduce(intersect,list(rownames(cd4_facs.data),
                                            rownames(cd14_facs.data),
                                                rownames(cd8_facs.data),
                                            rownames(nk_facs.data)))
pbmcs_reference <- as.matrix(cbind(cd4_facs.data[common_pbmcs_genes,],
                                   cd14_facs.data[common_pbmcs_genes,],
                                   cd8_facs.data[common_pbmcs_genes,],
                                   nk_facs.data[common_pbmcs_genes,]))
pbmcs_reference_labels <- c(rep('CD4',dim(cd4_facs.data)[2]),rep('CD14',dim(cd14_facs.data)[2]),
                            rep('CD8',dim(cd8_facs.data)[2]),rep('NK',dim(nk_facs.data)[2]))

```


## 4. Calculate rate and probability

```{r, message=FALSE}
# Train reference data
trainReference <- function(ref,pi.all=pi.all.g,mu=mu.g,a=a.all.g,sigma.all=sigma.all.g,
                           g.on.all=g.on.all.g,g.off.all=g.off.all.g,discrim_only=TRUE) {
  # Subset to genes present in both reference data and knowledge base
  common <- intersect(names(ref),rownames(pi.all))
  if (discrim_only) {
    common <- intersect(common,discrim.g)
  }
  N <- sum(ref)
  ref <- ref[common]
  pi.all2 <- pi.all[common,]
  a.all2 <- a[common]
  sigma.all2 <- sigma.all[common,]
  g.on.all2 <- g.on.all[common]
  g.off.all2 <- g.off.all[common]
  
  # Compute mixing probability for each component
  prob.exp <- pi.all2[,1]*dnbinom(ref,1,1/((N/(a.all2))+1))
  prob.ln1 <- sapply(common,function(j) 
    pi.all2[j,2]*sads::dpoilog(ref[j],mu[1]+g.off.all2[j]+log(N),sigma.all2[j,1]))
  prob.ln2 <- sapply(common,function(j) 
    (1-pi.all2[j,1]-pi.all2[j,2])*sads::dpoilog(ref[j],mu[2]+g.on.all2[j]+log(N),sigma.all2[j,2])) 
  return(data.frame(rate=ref/N,exp=prob.exp/(prob.exp+prob.ln1+prob.ln2),ln1=prob.ln1/(prob.exp+prob.ln1+prob.ln2),
                    ln2=prob.ln2/(prob.exp+prob.ln1+prob.ln2)))
}

# Train all reference
trainAllReference <- function(data,labels,pi.all=pi.all.g,mu=mu.g,a=a.all.g,sigma.all=sigma.all.g,
                              g.on.all=g.on.all.g,g.off.all=g.off.all.g,discrim_only=TRUE) {
  cell.types <- unique(labels)
  dataSums <- lapply(cell.types,function(c) rowSums(data[,labels==c]))
  d.list <- bplapply(dataSums,trainReference,pi.all=pi.all,mu=mu,a=a,sigma.all=sigma.all,
                     g.on.all=g.on.all,g.off.all=g.off.all,discrim_only=discrim_only)
  names(d.list) <- cell.types
  return(d.list)
}


pbmcs.d <- trainAllReference(pbmcs_reference,pbmcs_reference_labels,discrim_only=FALSE)
head(pbmcs.d[[1]])
```

## 5. getBarcode

```{r}
getBarcode <- function(d.list) {
  barcode <- sapply(d.list,function(x) 1-x[,2]-x[,3])      # ### Barcode is   ln2=prob.ln2/(prob.exp+prob.ln1+prob.ln2) if greater than 0
  barcode[barcode<0] <- 0
  colnames(barcode) <- names(d.list)
  rownames(barcode) <- rownames(d.list[[1]])
  return(barcode)
}

barcodes <- getBarcode(pbmcs.d)

```

## 6. Results

An example of a barcodes data frame

```{r}
class(barcodes)

barcodes<- as.data.frame(barcodes)

head(barcodes)
```

The length of expressed gene list for each cell type, if the probability of gene expression is great than 50% or  75%

```{r}
for (cut_off in c(0.5,0.75,0.95)) {
  cat("The following is the length of the expressed gene list for cell type CD4, CD14, CD8 and NK, if the probability of gene expression is great than ", cut_off , "\n")
  
  CD4_geneList <-rownames(barcodes)[barcodes$`CD4` > cut_off]
  cat("CD4:", length(CD4_geneList), "\n")
  write.table(CD4_geneList, file = paste0("CD4", cut_off, ".tsv"), sep = "\n", row.names = FALSE, col.names = FALSE, quote = FALSE)
  CD14_geneList <-rownames(barcodes)[barcodes$`CD14` > cut_off]
  cat("CD14:", length(CD14_geneList), "\n")
  write.table(CD14_geneList, file = paste0("CD14", cut_off, ".tsv"), sep = "\n", row.names = FALSE, col.names = FALSE, quote = FALSE)
  CD8_geneList <-rownames(barcodes)[barcodes$`CD8` > cut_off]
  cat("CD8:", length(CD8_geneList), "\n")
  write.table(CD8_geneList, file = paste0("CD8", cut_off, ".tsv"), sep = "\n", row.names = FALSE, col.names = FALSE, quote = FALSE)
  NK_geneList <-rownames(barcodes)[barcodes$`NK` > cut_off]
  cat("NK:", length(NK_geneList), "\n")
  write.table(NK_geneList, file = paste0("NK", cut_off, ".tsv"), sep = "\n", row.names = FALSE, col.names = FALSE, quote = FALSE)

}
```


## 7. Visualization

Venn diagram


## 8. Validation

compare results of the same cell types from different datasets e.g. cellRef vs. Hlca

compare results from different cell types. e.g. CD4 vs. CD8

Subsampling, Bootstrap Aggregation,Perturbation ( add 10% cells from other cell type)

compare with the known go term of the cell type.

## 9. Reporting
Summarize findings in a detailed report or manuscript with figures and statistical analyses.
