---
title: "Expression Probability (expProb)"
author: "Bingfang"
date: "2/12/2025"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Descriptions

The ultimate goal of this project is to identify a list of truly expressed genes in a specific cell type, which will be used for the functional analysis of biological processes and pathways in that cell type. This is in contrast to genes that are “differentially expressed” when comparing one cell type to other cell types or one condition to other conditions.

In previous works, I applied a method developed by Grakskit et al. (Statistics, 2023) to calculate the expression probability (expProb). The expProb method defines the expression distribution for each individual gene in three states:1) Null state – Modeled by an exponential component accounting for counts that were mostly zeros, consistent with practically no expression.2) Off state – Modeled by a log-normal component accounting for low counts, consistent with a nonzero background level of expression that is distinctly lower than the expressed state.3)On state – Modeled by a log-normal component accounting for the expressed state.I calculated the expression probability for each gene in each state for each lung cell type (see lung_geneExp_barcodeV4.Rmd).

To validate expProb, I calculated the expression probability of "on" state for NS-Forest genes (including marker genes and binary genes) in their targeted cell type—the cell type where they were identified as marker genes. However, about 25% of marker genes have "on" state expression probabilities lower than 75%, a artificial cutoff, in their targeted cell type (see lung_geneExp_barcodeV5.Rmd).

When calculating the expression probability of "on" state for differentially expressed (DE) genes. About 10% of DE genes have "on" state expression probabilities lower than 75% in the cell type where they are known to be expressed, see lung_geneExp_barcodeV6.Rmd).

It is a more stringent test to validate expProb using NS-Forest genes than using DE genes. Understanding why a subset of NS-Forest genes has predicted low expression probabilities could help us determine whether to use this expProb model to generate the expressed gene list for our knowledge graph and whether modification of model can improve the performance.

# 2. Data analysis plan
## 2.1 The purposes of this Rmarkdown

The question is: why does a subset of NS-Forest genes or differentially expressed genes have low expression probabilities in their target cell type? The hypothesis is that the parameters were trained on data collected before 2021. After re-training, the performance could be improved.


## 2.2 The rationales behind each analysis step

### 2.2.1 Analyze all prior and post measurements for the NS-forest markers

Analyze all prior and post measurements for the NS-forest markers with low expression probability (expProb, ln2) in the "on" state. Identify explanations for false negatives and explore opportunities for model improvement.

### 2.2.2 Calculate expression probability (expProb) using the updated parameter 'params_update_hlca.rda' from re-training

The old parameters (params_EM_81020.rda) were trained on data collected before 2021. After re-training, the performance could be improved.

### 2.2.3 Calculate recall and adjust precision when using NS-forest gene as true positive for validation

To investigate whether re-training improves recall without worsening precision.


 

# 3. Set environment


```{r, echo=TRUE,warning=FALSE,message=FALSE, results='hide'}
#install.packages("sads")  
library(sads)
#install.packages("BiocParallel")
library(BiocParallel)
library(Seurat)
library(org.Hs.eg.db)
#install.packages("Matrix")
library(Matrix)
# package for goenrichment
library(clusterProfiler)
library(AnnotationDbi)
# package for Venn diagram
library(gplots, lib.loc = "/usr/local/apps/R/4.4/site-library_4.4.2")
library(tidyverse)
# package for plot
install.packages("ggplot2")  
library(ggplot2)
# package for confusion matrix
install.packages('caret')
library(caret)
library(patchwork)

# Get gene-specific distributions
# Publication:https://academic.oup.com/biostatistics/article/23/4/1150/6622011
# https://github.com/igrabski/scRNAseq-cell-type
load('../params_update_hlca.rda') 
pi.all.g <- params[[1]]         # π-null,π-off, probability of null and off state.    
g.on.all.g <- params[[2]]       # μ0j shift from μ0 
g.off.all.g <- params[[3]]      # μ1j shift from μ1
a.all.g <- params[[4]]          # αj⁠
sigma.all.g <- params[[5]]      # σ0j,σ1j
mu.g<-c(-12.73153,-8.301349)    # global means for gene rates on a log scale, (μ0 , μ1)

```


# 4. Data 

### load HLCA core data from cellXgene

```{r, echo=TRUE,warning=FALSE,message=FALSE} 
## Read .rds file
#hlca = readRDS("/gpfs/gsfs10/users/xubr/cellXgene/scRNAseq-cell-type/data/b351804c-293e-4aeb-9c4c-043db67f4540_hlca_core.rds")

## Preview the dataset
#summary(hlca@assays$RNA@meta.features$feature_type)
#head(hlca@assays$RNA@meta.features$feature_name)
#table(hlca@meta.data$cell_type_ontology_term_id)
#table(hlca@meta.data$cell_type)
#table(hlca@meta.data$ann_finest_level)

```

### load data of NS-forest marker and binary marker genes, calculate prior probability.

```{r,echo=TRUE,warning=FALSE,message=FALSE} 
# The full table of gene symbol and Ensemble ID
gencode<-read.csv("../data/gencode.v47.annotation.csv")
gene_Ensemble <-gencode %>% select(c(`gene_name`, "ENSEMBLID"))

# load NS-forest marker gene and binary marker gene
#list.files("./data/")
binary_gene<-read.csv("../data/hlcacore_ann_finest_label_all_supplementary.csv", sep=",")
NSmarker_gene<-read.csv("../data/hlcacore_ann_finest_level.markers.csv", sep = ',')

# change format of NSmarker_gene
#NSmarker_gene$NSForest_markers[1]
NSmarker_gene$NSForest_markers <-gsub("]", "",NSmarker_gene$NSForest_markers)
NSmarker_gene$NSForest_markers <-gsub("\\[","",NSmarker_gene$NSForest_markers)
NSmarker_gene$NSForest_markers <-gsub(" ","",NSmarker_gene$NSForest_markers)
NSmarker_gene$NSForest_markers <-gsub("'","",NSmarker_gene$NSForest_markers)
NSmarker_gene <- NSmarker_gene %>%
  separate_rows(NSForest_markers, sep = ",")

# merge to obtain ENSEMBLID
NSmarker<-merge(NSmarker_gene,gene_Ensemble, by.x = "NSForest_markers", by.y = "gene_name", all.x = TRUE) 
geneList_bi<-merge(binary_gene,gene_Ensemble, by.x = "binary_genes", by.y = "gene_name", all.x = TRUE) 
#gencode[gencode$gene_name == "",] ## 4 NS-forest marker don't have EnsemblID


# generate a data frame of π-null,π-off, which is the probabilities of null and off states. 
pi_df<-as.data.frame(pi.all.g)
pi_df$`ENSEMBLID`<-rownames(pi_df)

# Obtain prior probability of expression for the NS-forest marker genes 
prior_marker<-merge(NSmarker, pi_df, by = "ENSEMBLID", all.x = TRUE) 
prior_marker$prior_prob_off <- prior_marker$V1 + prior_marker$V2
prior_marker$prior_prob_on <- 1-(prior_marker$V1 + prior_marker$V2)

# visualize prior probabilities for the NS-forest marker genes 
prior_marker_clean<-prior_marker %>% select(c("ENSEMBLID","NSForest_markers","clusterName","clusterSize","f_score","marker_count","binary_genes","V1","V2","prior_prob_off","prior_prob_on")) %>% drop_na()
#hist(as.numeric(prior_marker$prior_prob_on),main = "", xlab = "Prior probability of expression for the marker genes") 

# number of row remove due to NA
#dim(prior_marker)
#dim(prior_marker_clean)
# MTRNR2L8: no Ensemblid
# C1orf194: no Ensemblid
# CD68: no parameter

# Obtain prior probability of expression for the NS-forest binary genes 
prior_binary<-merge(geneList_bi, pi_df, by = "ENSEMBLID", all.x = TRUE) 
prior_binary$prior_prob_off <- prior_binary$V1 + prior_binary$V2
prior_binary$prior_prob_on <- 1-(prior_binary$V1 + prior_binary$V2)

# number of row remove due to NA
prior_binary_clean<- prior_binary%>%drop_na()
#dim(prior_binary)
#dim(prior_binary_clean)

# visualize prior probabilities for binary genes
#hist(as.numeric(prior_binary$prior_prob_on),main = "", xlab = "Prior probability of expression of the Binary genes")

```

Notes:

HLCA and cellRef are combined datasets. Will integration methods affect the proportion/rate (𝜆) of individual gene expression counts in total gene expression counts in a cell type?


# 5. Methods

## 5.1 Functions for calculating rate,three state probabilities

Functions from the 2022 Biostatistics paper


```{r, message=FALSE}

# Train reference data
trainReference <- function(ref,pi.all=pi.all.g,mu=mu.g,a=a.all.g,sigma.all=sigma.all.g,
                           g.on.all=g.on.all.g,g.off.all=g.off.all.g,discrim_only=FALSE) {
  # Subset to genes present in both reference data and knowledge base
  common <- intersect(names(ref),rownames(pi.all))
  if (discrim_only) {
    common <- intersect(common,discrim.g)
  }
  N <- sum(ref)
  ref <- ref[common]
  pi.all2 <- pi.all[common,]
  a.all2 <- a[common]
  sigma.all2 <- sigma.all[common,]
  g.on.all2 <- g.on.all[common]
  g.off.all2 <- g.off.all[common]
  
  # Compute mixing probability for each component
  prob.exp <- pi.all2[,1]*dnbinom(ref,1,1/((N/(a.all2))+1))
  prob.ln1 <- sapply(common,function(j) 
    pi.all2[j,2]*sads::dpoilog(ref[j],mu[1]+g.off.all2[j]+log(N),sigma.all2[j,1]))
  prob.ln2 <- sapply(common,function(j) 
    (1-pi.all2[j,1]-pi.all2[j,2])*sads::dpoilog(ref[j],mu[2]+g.on.all2[j]+log(N),sigma.all2[j,2])) 
  return(data.frame(rate=ref/N,exp=prob.exp/(prob.exp+prob.ln1+prob.ln2),ln1=prob.ln1/(prob.exp+prob.ln1+prob.ln2),ln2=prob.ln2/(prob.exp+prob.ln1+prob.ln2)))
}


# Train all reference
trainAllReference <- function(data,labels,pi.all=pi.all.g,mu=mu.g,a=a.all.g,sigma.all=sigma.all.g,
                              g.on.all=g.on.all.g,g.off.all=g.off.all.g,discrim_only=FALSE) {
  cell.types <- unique(labels)
  dataSums <- lapply(cell.types,function(c) rowSums(data[,labels==c]))

  d.list <- bplapply(dataSums,trainReference,pi.all=pi.all,mu=mu,a=a,sigma.all=sigma.all,
                     g.on.all=g.on.all,g.off.all=g.off.all,discrim_only=discrim_only)
  names(d.list) <- cell.types
  return(d.list)
}


# getBarcode of on state
getBarcode<-function(d.list) {
  barcode <- sapply(d.list,function(x) 1-x[,2]-x[,3])      # ### Barcode is   ln2=prob.ln2/(prob.exp+prob.ln1+prob.ln2) if greater than 0
  barcode[barcode<0] <- 0
  colnames(barcode) <- names(d.list)
  rownames(barcode) <- rownames(d.list[[1]])
  return(barcode)
}

# get probability of on and off state
getBarcode2 <- function(d.list) {
  barcode <- sapply(d.list,function(x) 1-x[,2])
  barcode[barcode<0] <- 0
  colnames(barcode) <- names(d.list)
  rownames(barcode) <- rownames(d.list[[1]])
  return(barcode)
}
```


## 5.2. Functions for calculating cell type-specific expression probability for lung datasets 

Functions to generate expression probability for cell types

```{r, message=FALSE}

## for cellRef, meta_feature is "celltype_level3_fullname"
## for hlca, meta_feature is "ann_finest_level"
## Run individual cell type, output expression probability of on state
generateCellType_barcode <- function(seurat_dataset,dataset_name, cellType, cut_off){
  # subset by cell type
  if(dataset_name=="hlca") {
  data_subset <- subset(seurat_dataset, subset = ann_finest_level == cellType)
  } else if(dataset_name=="cellRef"){
  data_subset <- subset(seurat_dataset, subset = celltype_level3_fullname == cellType)
  }
  matrix<- as.matrix(data_subset@assays$RNA@counts)

  labels<-c(rep(cellType,dim(matrix)[2]))
  
  # calculte rate and probability
  data.d <- trainAllReference(matrix,labels,discrim_only=FALSE)

  # get barcode and gene list
  barcodes <- getBarcode(data.d)
  geneList <-rownames(barcodes)[barcodes > cut_off]
  length(geneList)
  #write.table(geneList, file = paste0(cellType, cut_off, ".tsv"), sep = "\n", row.names = FALSE, col.names = FALSE, quote = FALSE)
  return(barcodes)
  
}

# Run individual cell type, return rate and three probabilities
generateCellType_rate <- function(seurat_dataset,dataset_name, cellType, cut_off){
  # subset by cell type
  if(dataset_name=="hlca") {
  data_subset <- subset(seurat_dataset, subset = ann_finest_level == cellType)
  } else if(dataset_name=="cellRef"){
  data_subset <- subset(seurat_dataset, subset = celltype_level3_fullname == cellType)
  }
  matrix<- as.matrix(data_subset@assays$RNA@counts)

  labels<-c(rep(cellType,dim(matrix)[2]))
  
  # calculte rate and probability
  data.d <- trainAllReference(matrix,labels,discrim_only=FALSE)

  # get barcode and gene list
  barcodes <- getBarcode(data.d)
  geneList <-rownames(barcodes)[barcodes > cut_off]
  length(geneList)
  #write.table(geneList, file = paste0(cellType, cut_off, ".tsv"), sep = "\n", row.names = FALSE, col.names = FALSE, quote = FALSE)
  return(data.d)
  
}


## Run multiple cell types, return expression probability of on state
runBatch<-function(seurat_dataset,dataset_name,cell_type_list,cut_off_list){
  dic ={}
  for(cellType in cell_type_list){
   
    for (cut_off in cut_off_list) {
      
      code<-generateCellType_barcode(seurat_dataset, dataset_name,cellType, cut_off)
      print(head(code))
      dic[[cellType]]<-as.data.frame(code) 
    }
  }
  return(dic)
}

## Run multiple cell types,return rate and three probabilities
runBatch_rate<-function(seurat_dataset,dataset_name,cell_type_list,cut_off_list){
  dic ={}
  for(cellType in cell_type_list){
   
    for (cut_off in cut_off_list) {
      
      code<-generateCellType_rate(seurat_dataset, dataset_name,cellType, cut_off)
      print(head(code))
      dic[[cellType]]<-as.data.frame(code) 
    }
  }
  return(dic)
}


## Check cell type-specific gene expression probability (ExpProb)
# output the expression probability for a gene in a specific cell type
checkGene <- function(id,sym, barcode_dic, marker, data_name){
  df_cell_type_gene<-data.frame(matrix(vector(), 0, 6,
                dimnames=list(c(), c("id", "sym", "barcode","marker", "cellType", "data"))),
                stringsAsFactors=F)
  for(i in 1:length(barcode_dic)){
    barcode<-barcode_dic[[i]][id,]
    df = data.frame(id=id,sym=sym, barcode=barcode, marker=marker,cellType=names(barcode_dic)[i], data=data_name)
    df_cell_type_gene <- rbind(df_cell_type_gene, df)
  }
  return(df_cell_type_gene)
}

# output the rate and three probability for a gene in a specific cell type
checkGene_rate <- function(id,sym, barcode_dic, marker, data_name){
  df_cell_type_gene<-data.frame(matrix(vector(), 0, 9,
                dimnames=list(c(), c("id", "sym", "marker", "cellType", "data", "rate","exp","ln1","ln2"))),
                stringsAsFactors=F)
  for(i in 1:length(barcode_dic)){
    barcode<-barcode_dic[[i]][id,]
    colnames(barcode)<-c("rate","exp","ln1","ln2")
    df = data.frame(id=id,sym=sym, marker=marker,cellType=names(barcode_dic)[i], data=data_name)
    df_rate<-cbind(df,barcode)
    df_cell_type_gene <- rbind(df_cell_type_gene, df_rate)
  }
  return(df_cell_type_gene)
}

```

# 6. Results


## 6.1 Calculate rate and three state probability for marker genes

```{r, message=FALSE}
## run all cell-type in HLCA
seurat_dataset<-hlca
dataset_name <-"hlca"
cell_type_list <- unlist(unique(hlca$ann_finest_level))
cut_off_list <- c(0.75)  # place holder for cut_off

#barcode_dic_detail<-runBatch_rate(seurat_dataset,dataset_name,cell_type_list,cut_off_list)
#saveRDS(barcode_dic_detail, file = "hlca_barcode_detail.rds")

barcode_dic_detail <- readRDS("../hlca_barcode_detail.rds")

# calculate rate and probability for NS-forest marker
data_name<-rep("hlca", nrow(prior_marker_clean))
df_hlca_gene_rate<-checkGene_rate(prior_marker_clean$ENSEMBLID,prior_marker_clean$NSForest_markers, barcode_dic=barcode_dic_detail, prior_marker_clean$clusterName, data_name)

# combine  prior and post probability for NS-forest marker
df_hlca_gene_rate<-df_hlca_gene_rate  %>% mutate(ENSEMBLID=id, 
                                                 clusterName=marker)
df_hlca_post_rate<-merge(df_hlca_gene_rate,prior_marker_clean,  by=c("ENSEMBLID","clusterName"),all.x = TRUE) %>% 
                   mutate(target=ifelse(clusterName == cellType, TRUE, FALSE)) 

marker_hlca_prior_poston_detail_post <- read.table("../processed_data/marker_hlca_prior_post_detail.tsv", sep = "\t", header = TRUE, stringsAsFactors = FALSE)


marker_hlca_prior_post_rate <-df_hlca_post_rate%>%  select(c("NSForest_markers","ENSEMBLID","clusterName","clusterSize","f_score","marker_count","data","cellType","V1","V2","prior_prob_off","prior_prob_on", "rate","exp","ln1","ln2","target"))
colnames(marker_hlca_prior_post_rate)<-c("NSForest_markers","ENSEMBLID","TargetCellType","clusterSize","f_score", "marker_count","data","dataCellSet","pi_null","pi_off","prior_prob_off","prior_prob_on", "rate","exp","ln1","ln2","targeted")
#write.table(marker_hlca_prior_post_rate, '../processed_data/marker_hlca_prior_post_detail.tsv', sep = "\t",row.names = FALSE,col.names = TRUE) 


```


## 6.2 Calculate update expProb for marker genes

```{r, message=FALSE}

#barcode_dic <- readRDS("hlca_barcode_update.rds")
```

Confirm from the author, expProb= ln2


## 6.3 Compare performance before and after retraining


Check the "on" state probability of NS-forest markers and calculate the confusion matrix. The following confusion matrix are based on assumption that marker genes expressed and only expressed in their targeted cell type. Since many markers expressed in more than their targeted cell type. FP counts were not correct. Therefore, recall is trustworthy, the precision is not correct.

### 6.3.1 Compare recall before after re-training


```{r, message=FALSE,warning=FALSE}

#### function for calculate recalls
calculate_recall <- function(x, df) {
  on_hlca_post <- df %>%
    mutate(expected_value = ifelse(targeted == "TRUE", 1, 0),
           predicted_value = ifelse(ln2 > x, 1, 0))

  common_levels <- union(levels(factor(on_hlca_post$predicted_value)),
                         levels(factor(on_hlca_post$expected_value)))

  on_hlca_post$predicted_value <- factor(on_hlca_post$predicted_value, levels = common_levels)
  on_hlca_post$expected_value <- factor(on_hlca_post$expected_value, levels = common_levels)

  # Creating confusion matrix
  on_matrix <- confusionMatrix(data = on_hlca_post$predicted_value, 
                               reference = on_hlca_post$expected_value, 
                               positive = "1")
  
  recall_value <- on_matrix$byClass['Sensitivity']
  return(recall_value)
}
```



```{r, message=FALSE,warning=FALSE}

#### before retraining
on_post <- read.table("../processed_data/marker_hlca_prior_post.tsv", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
on_post$ln2=on_post$expProb
cutoff<-seq(0.5,1, by=0.01)
# Call mapply with the function and the parameters
recall_pre <- mapply(calculate_recall, cutoff, MoreArgs = list(df = on_post))
data<-data_frame(cut0ff=cutoff, recall=unlist(recall_pre))
g_pre<-ggplot(data, aes(x = cutoff, y = recall)) +
  geom_line() +
  labs(title = "Before re-training",  # Add title
       x = "Cutoff Value",          # X-axis label
       y = "Recall (Sensitivity)") + # Y-axis label
  theme_minimal() +                  # Clean theme
  scale_y_continuous(limits = c(0, 1))


#### After retraining
on_detail_post <- read.table("../processed_data/marker_hlca_prior_post_detail.tsv", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
cutoff<-seq(0.5,1, by=0.01)
# Call mapply with the function and the parameters
recall_retrain <- mapply(calculate_recall, cutoff, MoreArgs = list(df = on_detail_post))
data<-data_frame(cut0ff=cutoff, recall=unlist(recall_retrain))
g_retrain<-ggplot(data, aes(x = cutoff, y = recall)) +
  geom_line()+
  labs(title = "After re-training",  # Add title
       x = "Cutoff Value",          # X-axis label
       y = "Recall (Sensitivity)") + # Y-axis label
  theme_minimal() +                  # Clean theme
  scale_y_continuous(limits = c(0, 1))

g_pre + g_retrain


```
### 6.3.2 Compare false position count before after re-training

```{r, message=FALSE,warning=FALSE}

#### function for calculate confusion matrix
calculate_confusion <- function(x, df) {
  on_hlca_post <- df %>%
    mutate(expected_value = ifelse(targeted == "TRUE", 1, 0),
           predicted_value = ifelse(ln2 > x, 1, 0))

  common_levels <- union(levels(factor(on_hlca_post$predicted_value)),
                         levels(factor(on_hlca_post$expected_value)))

  on_hlca_post$predicted_value <- factor(on_hlca_post$predicted_value, levels = common_levels)
  on_hlca_post$expected_value <- factor(on_hlca_post$expected_value, levels = common_levels)

  # Creating confusion matrix
  on_matrix <- confusionMatrix(data = on_hlca_post$predicted_value, 
                               reference = on_hlca_post$expected_value, 
                               positive = "1")
  print(cut_ff)
  print(on_matrix)
}


cutoff<-c(0.75,0.9,0.95)
#### before retraining
cm_pre <- mapply(calculate_confusion, cutoff, MoreArgs = list(df = on_post))
#### After retraining
cm_retrain <- mapply(calculate_confusion, cutoff, MoreArgs = list(df = on_detail_post))

```








## 8 Conclusion

The expProb model defines the expression of ~25% of NS-forest marker genes in the "off" state, likely because their expression levels are low. However, using a combined "on and off" state to identify expressed genes will result in a high false positive count and low precision.

Plans for next steps:


1.Make distribution plots for the markers with wrong expProb.
2.Check all prior and post parameter measure for the markers with wrong expProb, especially lamda.
3.Understand the training code and see if there is a room to modification.
4.Re-visit supplemental methods.
5.Set the target recall and precision. 
