---
title: "Data plan for geneExp"
author: "Bingfang"
date: "1/2/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Objective

Identify a list of true expressed genes of a specific cell type, which will use to for functional analysis of biological process/pathway of the cell type.

## 2. Set environment


```{r, echo=TRUE,warning=FALSE,message=FALSE, results='hide'}
#install.packages("sads")  
library(sads)
#install.packages("BiocParallel")
library(BiocParallel)
library(Seurat)
library(org.Hs.eg.db)
#install.packages("Matrix")
library(Matrix)
# package for goenrichment
library(clusterProfiler)
library(AnnotationDbi)
# package for Venn diagram
library(gplots, lib.loc = "/usr/local/apps/R/4.4/site-library_4.4.2")

# Get gene-specific distributions
# Publication:https://academic.oup.com/biostatistics/article/23/4/1150/6622011
# https://github.com/igrabski/scRNAseq-cell-type
load('params_EM_81020.rda') 
pi.all.g <- params[[1]]         # π-null,π-off, probabilty of null and off state.    
g.on.all.g <- params[[2]]       # σ1j⁠
g.off.all.g <- params[[3]]      # σ0j
a.all.g <- params[[4]]          # αj⁠
sigma.all.g <- params[[5]]      # μ0j,μ1j shift from (μ0 , μ1)
mu.g<-c(-12.73153,-8.301349)    # global means for gene rates on a log scale, (μ0 , μ1)

```

## 3. Data 

HLCA core data from cellXgene
cellRef data from cellXgene


```{r, echo=TRUE,warning=FALSE,message=FALSE} 
## Read .rds file
hlca = readRDS("/gpfs/gsfs10/users/xubr/cellXgene/scRNAseq-cell-type/data/b351804c-293e-4aeb-9c4c-043db67f4540_hlca_core.rds")
cellRef = readRDS("/gpfs/gsfs10/users/xubr/cellXgene/scRNAseq-cell-type/data/443f7fb8-2a27-47c3-98f6-6a603c7a294e_cellRef.rds")

## Preview the dataset
summary(hlca@assays$RNA@meta.features$feature_type)
#head(hlca@assays$RNA@meta.features$feature_name)
#table(hlca@assays$RNA@meta.features$feature_is_filtered)
table(hlca@meta.data$cell_type_ontology_term_id)
table(hlca@meta.data$cell_type)
table(hlca@meta.data$ann_finest_level)

summary(cellRef@assays$RNA@meta.features$feature_type)
#head(cellRef@assays$RNA@meta.features$feature_name)  ## lncRNA use Ensemble ID
#head(cellRef@assays$RNA@meta.features$origSymbol)    ## lncRNA use symbol
#table(cellRef@assays$RNA@meta.features$feature_is_filtered)
table(cellRef@meta.data$cell_type_ontology_term_id)
table(cellRef@meta.data$putative_CL_label)
table(cellRef@meta.data$celltype_level3_fullname)

```

Notes:
cellRef authors provided 47 cell ontology terms, which match to their 48 celltype_level3 clusters. They also provided putative cell ontology names.

HLCA authors provided 50 cell ontology terms with 50 cell type names, which partially match to their 61 finest-level clusters.

IncRNA ensembl ID may not be stable
e.g. ENSG00000238009 
Ensembl gene ENSG00000238009 is no longer in the database.It has been mapped to 1 current identifier  ENSG00000241860.


## 4 QC 

### 4.1 plot

```{r,warning=FALSE,message=FALSE}
## plot

jpeg(file="HLCA_nFeature.jpeg")
VlnPlot(hlca, features = c("nFeaturess_RNA_detected"))   #,raster=FALSE)
dev.off()

jpeg(file="cellRef_nFeature.jpeg")
VlnPlot(cellRef, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2)
dev.off()


```

### 4.2 Remove low quality cells

```{r,echo=TRUE,warning=FALSE,message=FALSE}
## filter
## need to convert to gene symbol to filter percent.mt < 5
## pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

hlca_filtered <- subset(hlca, subset = nFeaturess_RNA_detected > 200 & nFeaturess_RNA_detected < 2500)
cellRef_filtered <- subset(cellRef,subset = nFeature_RNA > 200 & nFeature_RNA < 2500)

## prior and after filter
hlca@assays$RNA@counts@Dim
hlca_filtered@assays$RNA@counts@Dim

cellRef$RNA@counts@Dim
cellRef_filtered@assays$RNA@counts@Dim
```


Notes:


HLCA and cellRef are combined datasets, Likely, they should not be filtered using nFeature.
High percentage cells will be filtered if using "nFeature_RNA > 200 & nFeature_RNA < 2500". 


## 5. Calculate rate,three state probabilities,and barcode

Functions from the 2022 Biostatistics paper


```{r, message=FALSE}

# Train reference data
trainReference <- function(ref,pi.all=pi.all.g,mu=mu.g,a=a.all.g,sigma.all=sigma.all.g,
                           g.on.all=g.on.all.g,g.off.all=g.off.all.g,discrim_only=FALSE) {
  # Subset to genes present in both reference data and knowledge base
  common <- intersect(names(ref),rownames(pi.all))
  if (discrim_only) {
    common <- intersect(common,discrim.g)
  }
  N <- sum(ref)
  ref <- ref[common]
  pi.all2 <- pi.all[common,]
  a.all2 <- a[common]
  sigma.all2 <- sigma.all[common,]
  g.on.all2 <- g.on.all[common]
  g.off.all2 <- g.off.all[common]
  
  # Compute mixing probability for each component
  prob.exp <- pi.all2[,1]*dnbinom(ref,1,1/((N/(a.all2))+1))
  prob.ln1 <- sapply(common,function(j) 
    pi.all2[j,2]*sads::dpoilog(ref[j],mu[1]+g.off.all2[j]+log(N),sigma.all2[j,1]))
  prob.ln2 <- sapply(common,function(j) 
    (1-pi.all2[j,1]-pi.all2[j,2])*sads::dpoilog(ref[j],mu[2]+g.on.all2[j]+log(N),sigma.all2[j,2])) 
  return(data.frame(rate=ref/N,exp=prob.exp/(prob.exp+prob.ln1+prob.ln2),ln1=prob.ln1/(prob.exp+prob.ln1+prob.ln2),ln2=prob.ln2/(prob.exp+prob.ln1+prob.ln2)))
}


# Train all reference
trainAllReference <- function(data,labels,pi.all=pi.all.g,mu=mu.g,a=a.all.g,sigma.all=sigma.all.g,
                              g.on.all=g.on.all.g,g.off.all=g.off.all.g,discrim_only=FALSE) {
  cell.types <- unique(labels)
  dataSums <- lapply(cell.types,function(c) rowSums(data[,labels==c]))

  d.list <- bplapply(dataSums,trainReference,pi.all=pi.all,mu=mu,a=a,sigma.all=sigma.all,
                     g.on.all=g.on.all,g.off.all=g.off.all,discrim_only=discrim_only)
  names(d.list) <- cell.types
  return(d.list)
}


# getBarcode
getBarcode<-function(d.list) {
  barcode <- sapply(d.list,function(x) 1-x[,2]-x[,3])      # ### Barcode is   ln2=prob.ln2/(prob.exp+prob.ln1+prob.ln2) if greater than 0
  barcode[barcode<0] <- 0
  colnames(barcode) <- names(d.list)
  rownames(barcode) <- rownames(d.list[[1]])
  return(barcode)
}
```


## 6. Output expressed gene lists and barcode dataframes for cell type

Function to un individual or a list of cell types

```{r, message=FALSE}

## for cellRef, meta_feature is "celltype_level3_fullname"
## for hlca, meta_feature is "ann_finest_level"
## Run individual cell type.
generateCellType_barcode <- function(seurat_dataset,dataset_name, cellType, cut_off){
  # subset by cell type
  if(dataset_name=="hlca") {
  data_subset <- subset(seurat_dataset, subset = ann_finest_level == cellType)
  } else if(dataset_name=="cellRef"){
  data_subset <- subset(seurat_dataset, subset = celltype_level3_fullname == cellType)
  }
  matrix<- as.matrix(data_subset@assays$RNA@counts)

  labels<-c(rep(cellType,dim(matrix)[2]))
  
  # calculte rate and probability
  data.d <- trainAllReference(matrix,labels,discrim_only=FALSE)

  # get barcode and gene list
  barcodes <- getBarcode(data.d)
  geneList <-rownames(barcodes)[barcodes > cut_off]
  length(geneList)
  write.table(geneList, file = paste0(cellType, cut_off, ".tsv"), sep = "\n", row.names = FALSE, col.names = FALSE, quote = FALSE)
  return(barcodes)
  
}


## Run multiple cell types
runBatch<-function(seurat_dataset,dataset_name,cell_type_list,cut_off_list){
  dic ={}
  for(cellType in cell_type_list){
   
    for (cut_off in cut_off_list) {
      
      code<-generateCellType_barcode(seurat_dataset, dataset_name,cellType, cut_off)
      print(head(code))
      dic[[cellType]]<-as.data.frame(code) 
    }
  }
  return(dic)
}

```

## 7. Validate the expression probabilities using know expressed genes e.g NS-forest markers


                     HLCA_NSForestMarkers	CellRef_NSForestMarkers
Alveolar macrophages ['MSR1', 'FABP4']	['MCEMP1', 'ACP5']
Alveolar Fibroblasts ['AOC3', 'LUM']	['SCN7A', 'ITGA8']

MSR1 ENSG00000038945 
FABP4 ENSG00000170323 
MCEMP1 ENSG00000183019
ACP5 ENSG00000102575

AOC3 ENSG00000131471 
LUM ENSG00000139329
SCN7A ENSG00000136546
ITGA8 ENSG00000077943 

```{r, message=FALSE,warning=FALSE}

## cell type name
#unique(cellRef$celltype_level3_fullname)
#unique(hlca$ann_finest_level)


## markers
marker<-c('Alveolar macrophage_HLCA_marker', 'Alveolar macrophage_HLCA_marker','Alveolar macrophage_cellRef_marker', 'Alveolar macrophage_cellRef_marker','Alveolar fibroblast 2_HLCA_marker', 'Alveolar fibroblast 2_HLCA_marker','Alveolar fibroblast 2_cellRef_marker', 'Alveolar fibroblast 2_cellRef_marker')
sym<-c('MSR1', 'FABP4','MCEMP1', 'ACP5','AOC3', 'LUM','SCN7A', 'ITGA8')
id <- c("ENSG00000038945","ENSG00000170323", "ENSG00000183019","ENSG00000102575","ENSG00000131471","ENSG00000139329","ENSG00000136546","ENSG00000077943")


## run 2 cell-type in HLCA
seurat_dataset<-hlca
dataset_name <-"hlca"
cell_type_list <- c("Alveolar macrophages","Alveolar fibroblasts", "AT1")
cut_off_list <- c(0.75,0.95)
barcode_dic2<-runBatch(seurat_dataset,dataset_name,cell_type_list,cut_off_list)
data_name<-rep("hlca", length(id))

checkGene <- function(id,sym, barcode_dic, marker, data_name){
  df_cell_type_gene<-data.frame(matrix(vector(), 0, 6,
                dimnames=list(c(), c("id", "sym", "barcode","marker", "cellType", "data"))),
                stringsAsFactors=F)
  for(i in 1:length(barcode_dic)){
    barcode<-barcode_dic[[i]][id,]
    df = data.frame(id=id,sym=sym, barcode=barcode, marker=marker,cellType=names(barcode_dic)[i], data=data_name)
    df_cell_type_gene <- rbind(df_cell_type_gene, df)
  }
  return(df_cell_type_gene)
}

df_hlca_gene<-checkGene(id,sym, barcode_dic=barcode_dic2, marker, data_name)
 
```



```{r, message=FALSE,warning=FALSE, results='hide'}
 ## run 2 cell-type in cellRef
seurat_dataset<- cellRef
dataset_name <- "cellRef"
cell_type_list <- c("Alveolar macrophage","Alveolar fibroblast 2","Alveolar type 1 cell")
cut_off_list <- c(0.75,0.95)
barcode_dic1<-runBatch(cellRef,dataset_name,cell_type_list,cut_off_list)

name<-rep("cellRef", length(id))
df_cellRef_gene<-checkGene(id,sym, barcode_dic=barcode_dic1, marker, data_name=name)


cell_type_gene<-rbind(df_cellRef_gene,df_hlca_gene)
cell_type_gene
write.table(cell_type_gene, 'cell_type_gene.tsv', sep = "\t",row.names = FALSE,col.names = TRUE)
```

## 8. Generate go term associated with cell type and compare with the known go term of the cell type.

extracted_object:
  id: 8e71666c-a590-4970-8794-fc0dae7a2760
  label: Type I pneumocyte
  biological_processes:
    - AUTO:gas%20exchange
    - GO:0003016
    - GO:0015671
    - GO:0015670
  cellular_component:
    - AUTO:alveolar%20surface
    - AUTO:pulmonary%20capillary
    - AUTO:air-blood%20barrier
named_entities:
  - id: AUTO:gas%20exchange
    label: gas exchange
  - id: GO:0003016
    label: respiratory gaseous exchange
  - id: GO:0015671
    label: oxygen transport
  - id: GO:0015670
    label: carbon dioxide transport
  - id: AUTO:alveolar%20surface
    label: alveolar surface
  - id: AUTO:pulmonary%20capillary
    label: pulmonary capillary
  - id: AUTO:air-blood%20barrier
    label: air-blood barrier
```{r}

library(clusterProfiler)
geneList <- read.csv("AT10.75.tsv")
gene<-unlist(geneList)
go <- enrichGO(gene=gene, OrgDb="org.Hs.eg.db",keyType = "ENSEMBL", ont="BP")
go_df <-as.data.frame(go)
write.table(go_df, 'AT1_075_go_term.tsv', sep = "\t",row.names = FALSE,col.names = TRUE)
```


GO:0003016
reactive oxygen species metabolic process	109	28	15.72	1.78	+	2.26E-03	3.57E-02
GO:0015671
oxygen transport	14	1	2.02	.50	-	7.08E-01	1.00E00
GO:0015670
carbon dioxide transport	14	3	2.02	1.49	+	4.41E-01	1.00E00


## 9. More Validation

### Compare results of the same cell types from different datasets e.g. cellRef vs. Hlca


```{r, warning=FALSE}

geneList_cellRef_Alveolar_fibroblast <- read.csv("Alveolar fibroblast 20.75.tsv")
geneList_hlca_Alveolar_fibroblast <- read.csv("Alveolar fibroblasts0.75.tsv")
venn(list(cellRef_Alveolar_fibroblast=geneList_cellRef_Alveolar_fibroblast,hlca_Alveolar_fibroblast=geneList_hlca_Alveolar_fibroblast))

```

### Compare results from different cell types. e.g. mucus_secreting cell vs. ionocyte.

```{r, warning=FALSE}
geneList_cellRef_Alveolar_fibroblast <- read.csv("Alveolar fibroblast 20.75.tsv")
geneList_cellRef_Alveolar_macrophages <- read.csv("Alveolar macrophages0.75.tsv")
geneList_cellRef_type1 <- read.csv("Alveolar type 1 cell0.75.tsv")
venn(list(cellRef_Alveolar_fibroblast=geneList_cellRef_Alveolar_fibroblast,cellRef_Alveolar_macrophages=geneList_cellRef_Alveolar_macrophages))

```

Generate DE gene list and compared with transitional DE list.



Subsampling, Bootstrap Aggregation,Perturbation ( add 10% cells from other cell type)

## 10. Reporting
Summarize findings in a detailed report or manuscript with figures and statistical analyses.

